---
title: "Final Project: Riya, Daniel, Andrew, Maya"
output: html_notebook
---


```{r}
library(dplyr)
library(ggplot2)
library(RCurl)
library(tidyr)

data<-getURL("https://media.githubusercontent.com/media/themarshallproject/crime-clearance-rate/refs/heads/main/crime_clearance_rate.csv")
crime<-read.csv(text=data)
```

#### Checkpoint 1

What is your imagined position on this dataset. Who has hired you, what are the overall goals?

```{r}
# Our team is positioned as analysts working for the US Government. We are tasked with supporting the FBI. The goal of our efforts is to help maximize crime clearance rate by identifying potential patterns in which crimes are more likely to be solved. We can do this by understanding trends and providing actionable insights that improve investigative efficiency and resource allocation across law enforcement agencies in the US.

```

What is the origin of the dataset. Who recorded the data? I promise you, it wasn't a user on kaggle.

```{r}
# This dataset originates from research conducted by The Marshall Project, which is a nonprofit organization dedicated to highlighitng the national urgency surrounding the US criminal justice system. They analyzed data from the FBI's unicorm crime reporting program to determine which types of crimes are most likley to be solved. Additionally, they examined the number of criems reported by agencies and the number of cases solved each year.

```

Explain what all of your remaining columns represent, and whether they are text, numeric, or categorical in nature. 

```{r}
# The columns ori and agency_name are text identifiers for each police agency. State is categorical, year is numeric, and “fips_state_county_code” and “number_of_months_missing” are numeric.

# The “new_” columns are numeric counts of reported crimes, and the corresponding “_cleared” columns are numeric counts of solved crimes.

# The dataset also includes numeric totals such as violent_crime, property_crime, total_crimes. Lastly, all “clearance_rate_” columns are numeric percentages showing the share of cases solved for each crime type and for crime overall.

``` 

## Variables Removed and Why

To prepare the dataset for our four research questions, we removed all
variables that did not directly relate to changes in clearance rates,
differences between states, or prediction/classification of overall
clearance rate.

### Removed Variables

**1. Identification variables (not useful for analysis)**
- `ori`, `agency_name`, `fips_state_county_code`
These identify agencies but do not contribute to crime patterns.

**2. Cleared-case count variables**
- All `cleared_*` columns
We use clearance *rates* instead of counts, so these are redundant and
introduce information leakage.

**3. Crime types not included in our research questions**
- `new_murder`, `new_manslaughter`, `new_rape`, `new_arson`,
  `new_total_crime`
Our focus is specifically on assault, robbery, burglary, theft, and motor
vehicle theft.

**4. Other clearance rate variables**
- All `clearance_rate_*` variables except `clearance_rate_total_crime`
These represent unrelated crime categories or alternate target variables.

### Variables Kept

We kept only variables relevant to our questions:

- `state`, `year`
- `new_assault`, `new_robbery`, `new_burglary`, `new_theft`,
  `new_motor_vehicle_theft`
- `clearance_rate_total_crime` (our main response variable)
#filtered dataset
```{r}
crime_filtered <- select(
  crime,
  state, year,
  new_assault, new_robbery, new_burglary, new_theft, new_motor_vehicle_theft,
  clearance_rate_total_crime
)
``` 
#### Checkpoint 2

How have clearance rates for violent crimes vs. property crimes changed over time in different states, and which states show the biggest improvement or decline? (Visualization)
Does New York differ from other states in its clearance rates for violent crime, burglary, or theft, and how do those differences change over time?
Can overall clearance rate be predicted using crime volume variables like number of assaults, robberies, burglaries, thefts, and motor vehicle thefts? (Linear Regression)
Can we classify whether a specific crime category (e.g., murder, robbery, burglary) will have a “high” vs. “low” clearance rate based on crime totals and state/year characteristics? (Logistic Regression or Random Forest)

```{r}
crime_ny_ca <- crime_filtered %>%
  drop_na(clearance_rate_total_crime) %>%
  filter(state %in% c("new york","california")) %>%
  filter(is.finite(clearance_rate_total_crime)) %>%
  group_by(state,year) %>%
  summarize(total_clearance_rate=sum(clearance_rate_total_crime)) 

ggplot(crime_ny_ca,aes(x=year,y=total_clearance_rate,color=factor(state),group=state)) + 
  geom_line(size=1) + 
  geom_point(size=2) +
  theme_minimal() +
  labs(x="Year",y="Clearance Rate",title="Crime Clearance Rate in New York vs California by Year",color="State") +
  theme(legend.position="bottom") +
  scale_y_continuous(labels=scales::comma,limits=c(5000,25000)) +
  scale_color_manual(values=c("#FFB900","#5773CC"))
```

```{r}
head(crime)

crime_clean<- crime %>% 
  select(clearance_rate_total_crime,
         new_assault, 
         new_robbery, 
         new_burglary, 
         new_theft, 
         new_motor_vehicle_theft) %>% 
  mutate_all(~ifelse(is.infinite(.)| is.nan(.), NA, .)) %>% 
  drop_na()

model<-lm(clearance_rate_total_crime ~
            new_assault+
            new_robbery+
            new_burglary+
            new_theft+
            new_motor_vehicle_theft,
          data=crime_clean)

summary(model)
```

# Written Response: Based on the linear regression model, the total number of different types of crimes does have statistically significant relationships with the overall clearance rate. Some crimes (such as assault and robbery) are associated with slightly higher clearance rates, while others (like burglary, theft, and motor vehicle theft) are associated with lower clearance rates.
However, the model’s R² value is extremely low (0.0003). This means that crime totals do not meaningfully predict how high a state’s clearance rate will be. The clearance rate is influenced by many factors outside of crime volume, including police resources, training, investigative practices, and state-level policies. Crime totals have statistically detectable effects, but they are not strong predictors of overall clearance rates. 

